Rapid-response Bayesian sky localization for electromagnetic follow-up of gravitational-wave candidates

### 2013-05-23 ms.tex
Remove some redundant definitions
Introduce u = cos(inclination)

A \ac{CBC} source is specified by a vector of extrinsic parameters describing its position and orientation and intrinsic parameters describing the physical properties of the binary components:
%
\begin{equation}
    \boldsymbol\theta = \begin{array}{l@{}l}
            \left[
            \begin{array}{c}
                \alpha \\
                \delta \\
                D_\mathrm{L} \\
                t_\oplus \\
                \iota \\
                \psi \\
                \phi_c \\
                \hdashline[1pt/1pt]
                m_1 \\
                m_2 \\
                \mathbf S_1 \\
                \mathbf S_2
            \end{array}
            \right] &
            \begin{array}{l}
                \left.
                \begin{array}{p{4cm}}
                    right ascension \\
                    declination \\
                    luminosity distance \\
                    arrival time at geocenter \\
                    inclination angle \\
                    polarization angle \\
                    coalescence phase \\
                    \hdashline[1pt/1pt]
                \end{array}
                \quad \right\} \textnormal{extrinsic parameters}, \boldsymbol\theta_\mathrm{ex} \\
                \left.
                \begin{array}{p{4cm}}
                    first component's mass \\
                    second component's mass \\
                    first component's spin \\
                    second component's spin
                \end{array}
                \quad \right\} \textnormal{intrinsic parameters}, \boldsymbol\theta_\mathrm{in}
            \end{array}
        \end{array}
\end{equation}
%
Binary eccentricity is omitted as an intrinisc parameter because tidal interaction of the progenitor stars~\cite{0004-637X-572-1-407} and later \ac{GW} emission~\cite{PhysRev.136.B1224} circularize the orbit long before the inspiral enters \ac{LIGO}'s frequency range of $\sim$10\nobreakdashes--1000~kHz. Tidal deformabilities of the \acp{NS} are omitted because the effect of tides on the inspiral would only be detectable by an Einstein Telescope-class \ac{GW} observatory~\cite{PhysRevD.81.123016}. Furthermore, in \ac{GW} detection efforts, especially those focused on \ac{NSNS} systems, the component spins $\mathbf{S}_1$ and $\mathbf{S}_2$ are often assumed to be aligned with the system's total angular momentum and condensed to a single scalar parameter $\chi$, or even neglected entirely, $\mathbf{S}_1 = \mathbf{S}_2 = 0$. In this work we will consider only zero spin.

Assuming circular orbits and no spin precession, we can write the \ac{GW} signal in each detector as a linear combination of two basis waveforms $h_0$ and $h_{\pi/2}$. For nonprecessing systems, $h = h_0$ and $h_{\pi/2}$ are approximately ``in quadrature'' in the same sense as the sine and cosine functions, being nearly orthogonal and out of phase by ${\pi/2}$ at all frequencies. In other words, $h_{\pi/2}(\omega) = i h_0(\omega)$ for all $\omega$. This linear combination isolates all dependence on the extrinsic parameters $\boldsymbol\theta_\mathrm{ex}$ into the coefficients and all dependence on the intrinsic parameters $\boldsymbol\theta_\mathrm{in}$ into the basis waveform,
%
\begin{eqnarray}\label{eq:full-signal-model}
	\fl h_i(\omega; \boldsymbol\theta) &=& e^{-i\omega t_\oplus}\frac{D_{1,i}}{D_\mathrm{L}}
	\left(\begin{array}{cc}
		F_{+,i}(\alpha, \delta, t_\oplus) \\
		F_{\times,i}(\alpha, \delta, t_\oplus)
	\end{array}\right)^\intercal
	\left(\begin{array}{cc}
		\cos{2\psi} & -\sin{2\psi} \\
		\sin{2\psi} & \cos{2\psi}
	\end{array}\right)
	\left(\begin{array}{cc}
		\frac{1}{2}\left(1+\cos^2{\iota}\right) & 0 \\
		0 & \cos{\iota}
	\end{array}\right) \nonumber\\\fl &\cdot&
	\left(\begin{array}{cc}
		\cos{2\phi_c} & -\sin{2\phi_c} \\
		\sin{2\phi_c} & \cos{2\phi_c}
	\end{array}\right)
	\left(\begin{array}{cc}
		1 \\
		i
	\end{array}\right) h(\omega; \boldsymbol\theta_\mathrm{in}).
\end{eqnarray}
%
This linear combination matches the conventions used in \cite{PhysRevD.83.084002}, and is the frequency-domain equivalent of their Equation~(2.8). The quantities $F_{+,i}$ and $F_{\times,i}$ are the dimensionless detector antenna factors, and depend on the orientation of detector $i$ as well as the sky location and sidereal time of the event, and are derived in~\cite{PhysRevD.63.042003}. The quantity $D_{1,i}$ is a fiducial horizon distance for detector $i$,
%
\begin{equation}\label{eq:horizon}
D_{1,i} = 1 / \sigma_i, \qquad {\sigma_i}^2 = \int_0^\infty \frac{\left|h(\omega; \mathbf \theta)\right|^2}{S_i(\omega)} \,\mathrm{d}\omega.
\end{equation}

@@

The choice of prior is open to one's astrophysical preconceptions, but during \ac{S6} when \ac{LIGO}'s Bayesian \ac{CBC} parameter estimation pipelines were pioneered, the prior was taken to be isotropic in sky location and binary orientation, and uniform in volume, arrival time, and the component masses~\cite{2013arXiv1304.1775T}. Put briefly, the prior was uniform in all of $(\alpha, \sin\delta, {D_\mathrm{L}}^3, t_\oplus, \cos\iota, \psi, \phi_c, m_1, m_2)$.

@@

\marginpar{Explain why none of these expressions depend on the coalescence phase $\phi_c$.}%
%
If we assume that the time and amplitude estimation errors are uncorrelated, then we can separate the likelihood into one factor that depends only on time and one that depends only on amplitude:
%
\begin{eqnarray*}
    \mathcal{L}(
        \hat{\tau}_1, \hat{\tau}_2, &\dots, \hat{\tau}_N,
        \hat{\rho}_1, \hat{\rho}_2, \dots, \hat{\rho}_N;
        \mathbf{n}, t_\oplus, D_\mathrm{L}, x, \psi) \\
    &= \mathcal{L}_\mathrm{TOA}(
        \hat{\tau}_1, \hat{\tau}_2, \dots, \hat{\tau}_N; \mathbf{n}, t_\oplus) \\
        &\times
        \mathcal{L}_\mathrm{SNR}(\hat{\rho}_1, \hat{\rho}_2, \dots, \hat{\rho}_N;
         \mathbf{n}, D_\mathrm{L}, x, \psi
    ).
\end{eqnarray*}
%
The \ac{TOA}-only factor depends only on the \ac{TOA} estimates $\hat{\tau}_1, \hat{\tau}_2, \dots, \hat{\tau}_N$, the unknown sky location $\mathbf{n}$ (or equivalently, the right ascension $\alpha$, declination $\delta$, and sidereal time), and the absolute \ac{TOA} at an arbitrary reference point (herein, the geocenter), $t_\oplus$, the fixed geographical locations of the detectors, $\mathbf{r}_i$, and the speed of light $c$.
%
\begin{equation}
    \mathcal{L}_\mathrm{TOA} \propto
    \exp \left[
            -\frac{1}{2} \sum_{i=1}^N \frac{\left(\hat{\tau}_i - \tau_i\right)^2}{\left(\hat{\rho_i} / \omega_{\mathrm{rms},i}\right)^2}
        \right]
    = \exp \left[
        -\frac{1}{2} \sum_{i=1}^N \frac{\left(\hat{\tau}_i + \mathbf{n} \cdot \mathbf{r}_i / c - t_\oplus \right)^2}{\left(\hat{\rho_i} / \omega_{\mathrm{rms},i}\right)^2}
    \right]
\end{equation}
%
The \ac{SNR}-only factor depends only on the \ac{SNR} estimates $\hat{\rho}_1, \hat{\rho}_2, \dots, \hat{\rho}_N$, but it depends on the unknown sky location, luminosity distance $D_\mathrm{L}$, cosine-inclination $x = \cos \iota$, and polarization angle $\psi$.
%
\begin{equation}
    \mathcal{L}_\mathrm{SNR} \propto \exp \left[
        -\frac{1}{2} \sum_{i=1}^N \left(\hat{\rho}_i - \rho_i(\mathbf{n}, D_\mathrm{L}, x, \psi) \right)^2
    \right]
\end{equation}
%
\marginpar{Cite \ac{GRB} prospects paper appendix; a very similar equation appears there. Is there somewhere else that has this formula? Shutz doesn't, and Finn \& Chernoff's notation is too heavy.}%
%
It is easily shown that
%
\begin{eqnarray*}
    {\rho_i}^2 &=
        \frac{1}{8} \left(\frac{D_{1,i}}{D_\mathrm{L}}\right)^2 \left[
            ( {F_{+,i}}^2 + {F_{\times,i}}^2 \right)
            \left(1 + 6 x^2 + x^4 \right) \\
            &+ \left(1 - x^2 \right) \left( {F_{+,i}}^2 - {F_{\times,i}}^2 \right) \cos {2\psi}
            + 2 \left( {F_{+,i}} {F_{\times,i}} ) \sin {2\psi}
        \right],
\end{eqnarray*}
%
where $F_{+,i}$ and $F_{\times,i}$ are the `$+$' and `$\times$' antenna factors of detector $i$, themselves functions of sky location $\mathbf{n}$.

>>>>

A \ac{CBC} source is specified by a vector of extrinsic parameters describing its position and orientation and intrinsic parameters describing the physical properties of the binary components:
%
\begin{equation}
    \boldsymbol\theta = \begin{array}{l@{}l}
            \left[
            \begin{array}{c}
                \alpha \\
                \delta \\
                D_\mathrm{L} \\
                t_\oplus \\
                u \\
                \psi \\
                \phi_c \\
                \hdashline[1pt/1pt]
                m_1 \\
                m_2 \\
                \mathbf S_1 \\
                \mathbf S_2
            \end{array}
            \right] &
            \begin{array}{l}
                \left.
                \begin{array}{p{4cm}}
                    right ascension \\
                    declination \\
                    luminosity distance \\
                    arrival time at geocenter \\
                    $\cos(\textrm{inclination angle})$ \\
                    polarization angle \\
                    coalescence phase \\
                    \hdashline[1pt/1pt]
                \end{array}
                \quad \right\} \textnormal{extrinsic parameters}, \boldsymbol\theta_\mathrm{ex} \\
                \left.
                \begin{array}{p{4cm}}
                    first component's mass \\
                    second component's mass \\
                    first component's spin \\
                    second component's spin
                \end{array}
                \quad \right\} \textnormal{intrinsic parameters}, \boldsymbol\theta_\mathrm{in}
            \end{array}
        \end{array}
\end{equation}
%
Binary eccentricity is omitted as an intrinisc parameter because tidal interaction of the progenitor stars~\cite{0004-637X-572-1-407} and later \ac{GW} emission~\cite{PhysRev.136.B1224} circularize the orbit long before the inspiral enters \ac{LIGO}'s frequency range of $\sim$10\nobreakdashes--1000~kHz. Tidal deformabilities of the \acp{NS} are omitted because the effect of tides on the inspiral would only be detectable by an Einstein Telescope-class \ac{GW} observatory~\cite{PhysRevD.81.123016}. Furthermore, in \ac{GW} detection efforts, especially those focused on \ac{NSNS} systems, the component spins $\mathbf{S}_1$ and $\mathbf{S}_2$ are often assumed to be aligned with the system's total angular momentum and condensed to a single scalar parameter $\chi$, or even neglected entirely, $\mathbf{S}_1 = \mathbf{S}_2 = 0$. In this work we will consider only zero spin.

Assuming circular orbits and no spin precession, we can write the \ac{GW} signal in each detector as a linear combination of two basis waveforms $h_0$ and $h_{\pi/2}$. For nonprecessing systems, $h = h_0$ and $h_{\pi/2}$ are approximately ``in quadrature'' in the same sense as the sine and cosine functions, being nearly orthogonal and out of phase by ${\pi/2}$ at all frequencies. In other words, $h_{\pi/2}(\omega) = i h_0(\omega)$ for all $\omega$. This linear combination isolates all dependence on the extrinsic parameters $\boldsymbol\theta_\mathrm{ex}$ into the coefficients and all dependence on the intrinsic parameters $\boldsymbol\theta_\mathrm{in}$ into the basis waveform,
%
\begin{eqnarray}\label{eq:full-signal-model}
	\fl h_i(\omega; \boldsymbol\theta) &=& e^{-i\omega t_\oplus}\frac{D_{1,i}}{D_\mathrm{L}}
	\left(\begin{array}{cc}
		F_{+,i}(\alpha, \delta, t_\oplus) \\
		F_{\times,i}(\alpha, \delta, t_\oplus)
	\end{array}\right)^\intercal
	\left(\begin{array}{cc}
		\cos{2\psi} & -\sin{2\psi} \\
		\sin{2\psi} & \cos{2\psi}
	\end{array}\right)
	\left(\begin{array}{cc}
		\frac{1}{2}\left(1+u^2\right) & 0 \\
		0 & u
	\end{array}\right) \nonumber\\\fl &\cdot&
	\left(\begin{array}{cc}
		\cos{2\phi_c} & -\sin{2\phi_c} \\
		\sin{2\phi_c} & \cos{2\phi_c}
	\end{array}\right)
	\left(\begin{array}{cc}
		1 \\
		i
	\end{array}\right) h(\omega; \boldsymbol\theta_\mathrm{in}).
\end{eqnarray}
%
This linear combination matches the conventions used in \cite{PhysRevD.83.084002}, and is the frequency-domain equivalent of their Equation~(2.8). The quantities $F_{+,i}$ and $F_{\times,i}$ are the dimensionless detector antenna factors, and depend on the orientation of detector $i$ as well as the sky location and sidereal time of the event, and are derived in~\cite{PhysRevD.63.042003}. The quantity $D_{1,i}$ is a fiducial horizon distance for detector $i$,
%
\begin{equation}\label{eq:horizon}
D_{1,i} = 1 / \sigma_i, \qquad {\sigma_i}^2 = \int_0^\infty \frac{\left|h(\omega; \mathbf \theta)\right|^2}{S_i(\omega)} \,\mathrm{d}\omega.
\end{equation}

@@

The choice of prior is open to one's astrophysical preconceptions, but during \ac{S6} when \ac{LIGO}'s Bayesian \ac{CBC} parameter estimation pipelines were pioneered, the prior was taken to be isotropic in sky location and binary orientation, and uniform in volume, arrival time, and the component masses~\cite{2013arXiv1304.1775T}. Put briefly, the prior was uniform in all of $(\alpha, \sin\delta, {D_\mathrm{L}}^3, t_\oplus, u, \psi, \phi_c, m_1, m_2)$.

@@

\marginpar{Explain why none of these expressions depend on the coalescence phase $\phi_c$.}%
%
If we assume that the time and amplitude estimation errors are uncorrelated, then we can separate the likelihood into one factor that depends only on time and one that depends only on amplitude:
%
\begin{eqnarray*}
    \mathcal{L}(
        \hat{\tau}_1, \hat{\tau}_2, &\dots, \hat{\tau}_N,
        \hat{\rho}_1, \hat{\rho}_2, \dots, \hat{\rho}_N;
        \mathbf{n}, t_\oplus, D_\mathrm{L}, x, \psi) \\
    &= \mathcal{L}_\mathrm{TOA}(
        \hat{\tau}_1, \hat{\tau}_2, \dots, \hat{\tau}_N; \mathbf{n}, t_\oplus) \\
        &\times
        \mathcal{L}_\mathrm{SNR}(\hat{\rho}_1, \hat{\rho}_2, \dots, \hat{\rho}_N;
         \mathbf{n}, D_\mathrm{L}, x, \psi
    ).
\end{eqnarray*}
%
The \ac{TOA}-only factor depends only on the \ac{TOA} estimates $\hat{\tau}_1, \hat{\tau}_2, \dots, \hat{\tau}_N$, the unknown sky location $\mathbf{n}$ (or equivalently, the right ascension $\alpha$, declination $\delta$, and sidereal time), and the absolute \ac{TOA} at an arbitrary reference point (herein, the geocenter), $t_\oplus$, the fixed geographical locations of the detectors, $\mathbf{r}_i$, and the speed of light $c$.
%
\begin{equation}
    \mathcal{L}_\mathrm{TOA} \propto
    \exp \left[
            -\frac{1}{2} \sum_{i=1}^N \frac{\left(\hat{\tau}_i - \tau_i\right)^2}{\left(\hat{\rho_i} / \omega_{\mathrm{rms},i}\right)^2}
        \right]
    = \exp \left[
        -\frac{1}{2} \sum_{i=1}^N \frac{\left(\hat{\tau}_i + \mathbf{n} \cdot \mathbf{r}_i / c - t_\oplus \right)^2}{\left(\hat{\rho_i} / \omega_{\mathrm{rms},i}\right)^2}
    \right]
\end{equation}
%
The \ac{SNR}-only factor depends only on the \ac{SNR} estimates $\hat{\rho}_1, \hat{\rho}_2, \dots, \hat{\rho}_N$, but it depends on the unknown sky location, luminosity distance $D_\mathrm{L}$, cosine-inclination $u$, and polarization angle $\psi$.
%
\begin{equation}
    \mathcal{L}_\mathrm{SNR} \propto \exp \left[
        -\frac{1}{2} \sum_{i=1}^N \left(\hat{\rho}_i - \rho_i(\mathbf{n}, D_\mathrm{L}, u, \psi) \right)^2
    \right]
\end{equation}
%
\marginpar{Cite \ac{GRB} prospects paper appendix; a very similar equation appears there. Is there somewhere else that has this formula? Shutz doesn't, and Finn \& Chernoff's notation is too heavy.}%
%
It is easily shown that
%
\begin{eqnarray*}
    {\rho_i}^2 &=
        \frac{1}{8} \left(\frac{D_{1,i}}{D_\mathrm{L}}\right)^2 \left\{
            ( {F_{+,i}}^2 + {F_{\times,i}}^2 \right)
            \left(1 + 6 u^2 + u^4 \right) \\
            &+ \left(1 - u^2 \right)^2 \left[\left( {F_{+,i}}^2 - {F_{\times,i}}^2 \right) \cos {2\psi}
            + 2 \left( {F_{+,i}} {F_{\times,i}} ) \sin {2\psi} \right]
        \right\}.
\end{eqnarray*}
